Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-21 submission/manuscript.Rmd
results edit 4

It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs. [@samantaActivationXistEvolutionarily2022].

Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4B), such as _Tktl2_ and _Esx1_. We the compared the known functions of germline genes dysregulated only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared). Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, male-specifc DEGs had mitochondrial and cell signaling gene functions (GO:0070585 protein localization to mitochondrion). Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 meiotic nuclear division) and egg-specific functions (GO:0007292 female gamete generation).

The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D). This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated suppression of germline genes. Female sensitivity could be due to improper X chromosome inactivation (XCI) in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, both shared and female-specific germline DEGs were not biased towards the X chromosome and the majority of female DEGs instead lie on autosomes (Figure 4G). Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of potential defects in XCI.

>>>>

It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. Sex is particularly pertinent in the case of KDM5C because it lies on the X chromosome and partially escapes X chromsome inactivation (XCI), resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. We therefore explored the impact of chromosomal sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs. [@samantaActivationXistEvolutionarily2022].

Homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs, such as _Tktl2_ and _Esx1_ (Figure 4B). We compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared). Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 protein localization to mitochondrion). Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 meiotic nuclear division) and egg-specific functions (GO:0007292 female gamete generation).

The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D). This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression. Female sensitivity could be due to improper XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, female germline DEGs were not biased towards the X chromosome and had a a similar overall proportion of X chromosome DEGs compared to males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G). For both males and females, the majority of germline DEGs instead lie on autosomes (Figure 4G). Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of potential defects in XCI.
